Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-15 paper/eLife_original_submission.tex
add eLife digest answers

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk,saira2013sequence}.
These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (\FIG{viruspopulations}A) and to particles capable of inducing expression of a single viral protein (\FIG{viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (\FIG{viruspopulations}A,B).
We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (\FIG{viruspopulations}C).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry data for  \FIG{viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
\end{figure}

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, performing two slightly different variants of the experiment for the 8-hour timepoint.
For most of the samples, we replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
However, for the second 8-hour sample (which we denote as ``8hr-2'' in the figures), we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
The rationale for including a sample without a media change was to determine the importance of synchronicity of the timing of infection as discussed later in this subsection.

We recovered between 3,000 and 4,000 cells for each sample (\FIG{cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
\FIGSUPP[cells]{viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
}
\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
\end{figure}

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
\FIG{viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and \FIG{viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, likely reflecting co-infection.

\begin{figure}[t!]
\centerline{\includegraphics[width=0.82\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
This plot illustrates the process for the 10-hour sample, see \FIGSUPP[viralbarcodes]{thresholds} for similar plots for other samples.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
\FIGSUPP[viralbarcodes]{cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
}

\figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
The number of viral barcodes called for each sample and gene segment. 
Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
For calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
}{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}

\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
As is shown in \FIG{viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
We called the threshold at this point of maximum purity (dotted green line).
For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as \FIG{viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (\FIG{cells}B).
}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}

\figsupp[Cumulative distributions of viral mRNA per cell and Gini coefficients.\label{figsupp:cumulfracflu}]{
The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
For instance, the plot for the 8hrs sample shows that $\sim$50\% of all viral mRNA is derived from $\sim$8\% of the infected cells.
The facet titles above each plot also give the Gini coefficient~\citep{gini1921measurement} that calculates the heterogeneity in the distribution of viral mRNA among infected cells.
Gini coefficients of 0 indicate a perfectly even distribution across cells, and Gini coefficients of 1 indicate a maximally skewed distribution.
}{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}

\figsupp[Synchronization of infection does not greatly affect heterogeneity. \label{figsupp:synchronize}]
{\jdbcomment{Alistair, can you add caption}
}{\jdbcomment{Alistair, can you add add figure}}

\figsupp[Effects of infectious dose or coinfection state. \label{figsupp:flowdose}]
{
{\bf (A)}
Flow cytometry analysis of expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
While a higher dose leads to more cells expressing high amounts of viral protein, it does not greatly increase the amount of viral protein in either the low-expressing or high-expressing cells. 
Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the proportions of cells that fall in different parts of the highly heterogeneous distribution.
{\bf (B)}
Cells were co-infected with a mix of wild-type virus and pseudovirus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment at an MOI of 0.1 for each virus. 
At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry for these proteins and GFP.
Cells could be annotated as infected by virions of the same type (wild-type infection indicated by presence of HA, or pseudovirus infection indicated by the presence of GFP) or both types of virions (indicated by presence of HA and GFP).
Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
Replicates shown.
}{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
\end{figure}

We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C, \FIGSUPP[viralbarcodes]{thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
\FIG{viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and may miss some true low-level infections.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (\FIG{viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in \FIG{viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (\FIG{viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes]{cumulfracflu}).
One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
\FIGSUPP[viralbarcodes]{cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
The Gini coefficients are $\ge$0.64 for all samples.
As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.

One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (\FIG{viralbarcodes}E, \FIGSUPP[viralbarcodes]{cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
As shown in \FIGSUPP[viralbarcodes]{synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since \FIG{viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.

Notably, \FIG{viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (\FIGSUPP[viralbarcodes]{flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (\FIGSUPP[viralbarcodes]{flowdose}).

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.83\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
Plots for individual samples are in \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
The trends are similar if we look only at the 10-hour sample (\FIGSUPP[fluburdenbyflugene]{10hrfluburdenbyflugene}) or exclude known co-infected cells (\FIGSUPP[fluburdenbyflugene]{fluburdenbyflugene_nocoinfection}).
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}

\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}

\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in \FIG{fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}

\figsupp[Like panel (A), but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburden_nocoinfection}]{
The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, excluding cells that were annotated as coinfected based on the presence of both viral barcodes.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_no_coinfect.pdf}}

\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}

>>>>

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk,saira2013sequence}.
These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (\FIG{viruspopulations}A) and to particles capable of inducing expression of a single viral protein (\FIG{viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
We confirmed that our viral stocks induced less interferon than a stock propagated at higher MOI (\FIG{viruspopulations}C).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
Error bars $\pm$ S.D., n=6 (qPCR replicates). 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
Error bars $\pm$ S.D., n=3.
Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
}
\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
Full flow cytometry date for \FIG{viruspopulations}B. 
A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
{\bf (A)} Uninfected gating control.
{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
We recovered between 3,000 and 4,000 cells for each sample (\FIG{cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B).

\begin{figure}
\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the amount of viral mRNA per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
The number of cellular and viral mRNAs detected for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
}
\end{figure}

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
\FIG{viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and \FIG{viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection.

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a cutoff for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
We fit a curve to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
See the Methods for details.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
}
\figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
The number of viral barcodes called for each sample and gene segment. 
Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
For calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
}{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
}{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
\end{figure}

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
\FIG{viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in \FIG{viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (\FIG{viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes}]{cumulfracflu}).
Notably, co-infected cells do not express dramatically more viral mRNA than other cells (\FIG{viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.

@@

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
{\bf (C)}
The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
Points are colored as in \FIG{fluburdenbyflugene}A.
}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
We do not show statistical analyses for other samples, as the number of infected cells is too low.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}
